\specialchapt{ACKNOWLEDGEMENTS}

I would like to acknowledge the phenomenal mentorship provided by Volker Brendel and Amy Toth.
Working through many tough research problems over the years has been a bittersweet experience, and I treasure the critical thinking skills and healthy skepticism that they have passed on to me along the way.
If I have any success in my future scientific career it will be due in no small part to their instruction.

I express thanks to the other outstanding scientists on my graduate committee: Karin Dorman, Xiaoqui Huang, and Jonathan Wendel.
Their invaluable feedback and generosity with their time has been greatly appreciated, and I count myself fortunate for their support.

I am greatly indebted to Trish Stauble, not only for her infectious optimism and enthusiasm, but also for her indefatigable support.
She has been extremely generous with her time and helpful navigating paper work, logistics, and graduate school life in general.
The BCB ``First Thursday" dinners, many of which she generously hosted at her own home, will not be soon forgotten.
Trish's impact is unparallelled in so many ways.

My studies and work greatly benefited from discussions and feedback from my Brendel Group colleagues:
Taylor Raborn, Chun-Yuan Huang, Saranya Sankaranarayanan, Vineela Gangalapudi, Murat {\"O}zt{\"u}rk, and Brad Bowser.
I also acknowledge my Toth Lab colleagues who made my summer 2015 ``field sabbatical" a wonderful experience, and especially Ali Berens who has been an especially good colleague and friend throughout our graduate training.

My deepest gratitude is reserved for my family, and most especially the love of my life, Stephanie, without whose encouragement, support, counsel, and selflessness I would not be who I am today.
She is the yin to my yang, my superior in many ways, and my best friend forever.
This work is dedicated to her.
